Concurrent Programming in Harmony

### 2022-01-04 book/paper.tex
More on behaviors

\begin{comment}
The framework in \autoref{fig:csbarebones} works well for thoroughly
testing mutual exclusion and progress in critical sections, but depends
on \texttt{countLabel}, an unusual operator specific to Harmony that is mostly
useful for testing mutual exclusion.
It turns out that we do not need labels to express this.
\autoref{fig:cslock} tests mutual exclusion and progress for
critical sections implemented using locks.
The test uses a simple atomic counter that is incremented before entering
the critical section and decremented after leaving it.
Mutual exclusion holds if the counter never goes over 1.
\end{comment}

As with critical sections---when testing a concurrent data structure---we
need a specification.
For example, \autoref{fig:queues}(a) shows a sequential specification of a
queue in Harmony.
First, we can check if the queue implementation
in \autoref{fig:queueconc} meets the sequential
queue specification in \autoref{fig:queues}(a).
To check if the queue implementation meets the specification,
we need to see if any sequence of queue operations in the
implementation matches a corresponding sequence in the
specification.
We say that the implementation and the specification
have the same \emph{behavior}.
%
\autoref{fig:qtestseq} presents a test program that does
exactly this, for sequences of up to <{NOPS}>
queue operations.
It maintains two queues:
\begin{itemize}
\item[] <{specq}>: the queue of the specification;
\item[] <{implq}>: the queue of the implementation.
\end{itemize}
For each operation, the code first chooses whether to
perform a <{get}> or <{put}> operation.
In the case of a <{put}> operation, the code also
chooses which value to append to the queue.
All operations are performed on both the queue implementation
and the queue specification.
In the case of <{get}>, the results of the operation
on both the implementation and specification are checked against one another.

\emph{Test programs themselves should be tested}.
Just because a test program works with a particular implementation
does not mean the implementation is correct---it may be that the implementation
is incorrect but the test program does not have enough coverage to find any
bugs in the implementation.
So, run a test program like this with a variety of queue implementations that
have known bugs in them and make sure that the test program finds them.
\begin{comment}
In the case of testing mutual exclusion with atomic counters, you will want
to see if you can use this method to find the problems in \autoref{fig:naivelock},
\autoref{fig:naiveflags},
and \autoref{fig:naiveturn}.
\end{comment}
Conversely, a test program may be broken in that it finds bugs that do not exist.
In my experience, it is often harder to implement the test program than
the algorithm that the test program tests.

\begin{comment}
So, in this case,
you also want to check if an atomic counter solution works with known
correct solutions such as \autoref{fig:peterson} and \autoref{fig:tas}
if possible.
\end{comment}

@@

\section*{Exercises}
\begin{problems}
\item Write a Harmony program that checks if
\autoref{fig:linkedlist} satisfies the specification
of \autoref{fig:setobj} \emph{sequentially}.
\item Write a Harmony program that checks if
\autoref{fig:linkedlist} satisfies the specification
of \autoref{fig:setobj} \emph{concurrently}.
% \item Extend \autoref{fig:linkedlist} with linearization points and
% check if the implementation is linearizable.
% \item Check if the extended queue implementation of \autoref{ex:qremove}
% is linearizable.
\item Rewrite \autoref{fig:qtestseq} so it only imports \texttt{queue}
and runs <{NOPS}> nondeterministically chosen operations against
it (similar in style to \autoref{fig:qtestpar} but without threads).
Then use behaviors to check that \autoref{fig:queueconc} and
\autoref{fig:queueseq} are correct sequential implementations of
the queue.  Check your test program by also trying it on one or two
buggy queue implementations.
\end{problems}

>>>>

As with critical sections---when testing a concurrent data structure---we
need a specification.
For example, \autoref{fig:queues}(a) shows a sequential specification of a
queue in Harmony.
First, we can check if the queue implementation
in \autoref{fig:queueconc} meets the sequential
queue specification in \autoref{fig:queues}(a).
To check if the queue implementation meets the specification,
we need to see if any sequence of queue operations in the
implementation matches a corresponding sequence in the
specification.
We say that the implementation and the specification
have the same \emph{behaviors} or are
\emph{behaviorally equivalent}.

Behaviors say something about how we got to a state.
The same state can be reaching by multiple behaviors, and the behaviors
are often an integral part of whether a program is correct or not.
Just because a state satisfies some invariant---however important---does
not mean that the state is valid given the sequence of operations.
For example, a state in which the queue is empty is certainly a valid state in
its own right, but if the last operation to get there was an enqueue operation,
there must be a bug in the program.
It can therefore be important to capture the behaviors.
We could store behaviors in the state itself by adding what is
known as a \emph{history variable} that keeps track of all the operations.
While this can be useful for correctness proofs, for model checking this approach
presents a problem: introducing this
additional state can lead to state explosion or even turn a finite model
(a model with a finite number of states) into an infinite one.
We therefore use a different approach: composing an implementation with
its specification to ensure that accept the same behaviors.

\autoref{fig:qtestseq} presents a test program that does
exactly this, for sequences of up to <{NOPS}>
queue operations.
It maintains two queues:
\begin{itemize}
\item[] <{specq}>: the queue of the specification;
\item[] <{implq}>: the queue of the implementation.
\end{itemize}
For each operation, the code first chooses whether to
perform a <{get}> or <{put}> operation.
In the case of a <{put}> operation, the code also
chooses which value to append to the queue.
All operations are performed on both the queue implementation
and the queue specification.
In the case of <{get}>, the results of the operation
on both the implementation and specification are checked against one another.

\emph{Test programs themselves should be tested}.
Just because a test program works with a particular implementation
does not mean the implementation is correct---it may be that the implementation
is incorrect but the test program does not have enough coverage to find any
bugs in the implementation.
So, run a test program like this with a variety of queue implementations that
have known bugs in them and make sure that the test program finds them.
Conversely, a test program may be broken in that it finds bugs that do not exist.
In my experience, it is often harder to implement the test program than
the algorithm that the test program tests.

@@

\section*{Exercises}
\begin{problems}
\item \autoref{fig:lockspec} shows a specification of a lock.  Write a program
that checks the behaviors of lock implementations such as \autoref{fig:tas2}
and \autoref{fig:ticket}.  That is, it should not rely on assertions such as
in \autoref{fig:cs}.
\item Write a Harmony program that checks if
\autoref{fig:linkedlist} satisfies the specification
of \autoref{fig:setobj} \emph{sequentially}.
\item Write a Harmony program that checks if
\autoref{fig:linkedlist} satisfies the specification
of \autoref{fig:setobj} \emph{concurrently}.
% \item Extend \autoref{fig:linkedlist} with linearization points and
% check if the implementation is linearizable.
% \item Check if the extended queue implementation of \autoref{ex:qremove}
% is linearizable.
\item Rewrite \autoref{fig:qtestseq} so it only imports \texttt{queue}
and runs <{NOPS}> nondeterministically chosen operations against
it (similar in style to \autoref{fig:qtestpar} but without threads).
Then use behaviors to check that \autoref{fig:queueconc} and
\autoref{fig:queueseq} are correct sequential implementations of
the queue.  Check your test program by also trying it on one or two
buggy queue implementations.
\end{problems}
